---
title: " Boye Kinzig Stream Parasites "
author: "Annemie Doliwa and Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all objects before starting
knitr::opts_chunk$set(echo = TRUE)
```


#### load required packages

```{r packages, message=FALSE}
library(here)
library(tidyverse)
library(readxl)
library(vegan)
library(RColorBrewer)
```

# preliminary steps

### variables and functions

```{r}
input <- "full_table_Boye21_AD_OpenRefine05.10..xlsx"
metadata_file <- "SFB_BasicData_Classification_Boye & Kinzig_2022-02-06_BoyeOnly.xlsx"
output <- str_replace(input, "..xlsx", "_reduced_taxonomy.tsv")
annotated_table <- str_replace(output, ".tsv", ".xlsx")
eukaryotic_rarefied_file <- str_replace(input, "..xlsx", "_eukaryotic_rarefied.tsv")
parasitic_rarefied_file <- str_replace(input, "..xlsx", "_parasitic_rarefied.tsv")
seed <- 123
n_subsamplings <- 100
cbPalette <- c("#000000", "#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
load_raw_occurrence_data <- function(filename){
    here::here("data", filename) %>%
        readxl::read_xlsx(na = c("", "NA")) %>%
            select(-sequences) %>%
            mutate(across(.cols = -c(seqid, starts_with("taxonomy")),
                          .fns = as.numeric))
}

extract_and_format_metadata <- function(filename){
    here::here("data", filename) %>%
        readxl::read_xlsx(skip = 2, n_max = 22) %>%
            slice(3:23) %>%
            rename(site_name = characteristics,
                   stream_name = `...2`,
                   years_since_restauration = `...4`,
                   GSG_CRC_class = `...7`,
                   years_since_sewage = `...9`,
                   years_since_sewage_class = `...10`,
                   years_since_sewage_CRC_class = `...11`,
                   conductivity_CRC_class = `...13`,
                   dissolved_oxygen = `dissolved oxygen`,
                   dissolved_oxygen_CRC_class = `...15`,
                   sulfate = `sulfate NEW`,
                   sulfate_CRC_class = `...17`,
                   nitrate_CRC_class = `...19`,
                   overall_mean = `overall\r\nmean`) %>%
            select(-c(`...5`, DOC, `...21`, `...23`)) %>%
            mutate(across(everything(), ~ str_replace(., ".* (not assessed|never sewer) .*", "NA")),
                   across(everything(), ~ na_if(., "NA")),
                   years_since_sewage_class = str_remove(years_since_sewage_class, " *[Yy]ears?"),
                   restoration = str_replace(restoration, ".* reference .*", "-Inf"),
                   sewagefree = replace_na(sewagefree, -Inf),
                   years_since_restauration = replace_na(years_since_restauration, Inf),
                   years_since_sewage = replace_na(years_since_sewage, Inf),
                   across(c(ends_with("_class"), GSG), ~ as.factor(.)),
                   across(c(where(is.character), -ends_with("_name")), ~ as.numeric(.)))
}


. %>%
    rename(supergroup = `taxonomy 2`) %>%
    mutate(supergroup = str_remove(supergroup, " *\\([0-9]+\\)")) %>%
    select(-seqid, -starts_with("taxonomy")) %>%
    rowwise() %>%
    mutate(reads = sum(c_across(-supergroup)), .keep = "unused") %>%
    ungroup() %>%
    count(supergroup, wt = reads, name = "reads", sort = TRUE) %>%
    mutate(percentage = 100 * reads / sum(reads),
           cumulative_perc = cumsum(percentage)) -> list_supergroups
```


## extract metadata

The metadata table is formatted for human readers, not for R. Some leg
work is required to make the metadata usable.

```{r}
extract_and_format_metadata(metadata_file) -> metadata
```


## extract a list of unique taxa

That shorter list will then be annotated to identify parasitic taxa.

```{r}
load_raw_occurrence_data(input) %>%
        select(starts_with("taxonomy")) %>%
        mutate(across(everything(), ~ str_remove(., " *\\([0-9]+\\)"))) %>%
        arrange(across(everything())) %>%
        distinct() %>%
        write_tsv(here::here("data", output))
```


## extract a list of parasitic taxa

Annemie and Micah have annotate the table.

```{r}
here::here("data", annotated_table) %>%
    readxl::read_xlsx(skip = 1) %>%
        select(ParasiticTax) %>%
        filter(! is.na(ParasiticTax)) %>%
        distinct() %>%
        pull() -> known_parasites

known_parasites
```

There are `r length(known_parasites)` taxa marked at parasites, at
different taxonomic levels.


## checks and filtering

here we go!

```{r}
load_raw_occurrence_data(input) -> raw_table
```

how many reads per sample?

```{r}
raw_table %>%
    select(-starts_with("taxonomy")) %>%
    pivot_longer(cols = -seqid,
                 names_to = "samples",
                 values_to = "reads") %>%
    count(samples, wt = reads, name = "reads") %>%
    pull(reads) -> sample_sizes

sample_sizes %>%
    summary()

sample_sizes %>%
    min() -> smallest_sample

rm(sample_sizes)
```

how many reads per domain?

```{r}
taxonomy_columns_to_discard <- paste(c("taxonomy"), 2:8, sep = " ")

raw_table %>%
    select(-taxonomy_columns_to_discard) %>%
    rename(domain = `taxonomy 1`) %>%
    mutate(domain = str_remove(domain, " *\\([0-9]+\\)")) %>%
    pivot_longer(cols = -c(seqid, domain),
                 names_to = "samples",
                 values_to = "reads") %>%
    count(domain, wt = reads, name = "reads") %>%
    mutate(percentage = 100 * reads / sum(reads))

rm(taxonomy_columns_to_discard)
```

Note that there are no unassigned reads, which is a bit surprising,
even for 18S V9 reads. It is safe to exclude all reads assigned to
Bacteria.


# Eukaryota

Exclude bacteria:


```{r}
raw_table %>%
    filter(str_detect(`taxonomy 1`, "Bacteria", negate = TRUE)) -> eukaryotic_table

rm(raw_table)
```


## list supergroups

```{r}
eukaryotic_table %>%
    list_supergroups
```

The dataset is dominated by Opisthokonta, Stramenopiles and
Archaeplastida.


## rarefaction (random subsampling)

### local functions

Prepare to remove empty OTUs from the final rarefied table:

```{r}
. %>%
    t() %>%
    as.data.frame() %>%
    as_tibble() -> transpose_back_to_tibble

. %>%
    rowwise() %>%
    mutate(reads = sum(c_across(-seqid)), .keep = "all") %>%
    ungroup() %>%
    filter(reads > 0) %>%
    select(-reads) -> remove_empty_OTUs
```


### make a transposed table

Remove columns that are not samples:

```{r}
eukaryotic_table %>%
    select(-seqid, -starts_with("taxonomy")) %>%
    t() -> eukaryotic_table_transposed

eukaryotic_table %>%
    select(seqid) -> OTU_ids
```


### randomly subsample the table

Randomly subsample the table, so all samples have the same number of
reads. Repeat the process `r n_subsamplings` times to make sure that
the final profile is as close as possible to the initial
distribution. Use a fix seed to make the process 100% repeatable. That
step can take several minutes to run.

```{r}
set.seed(seed)
matrix1 <- vegan::rrarefy(eukaryotic_table_transposed, smallest_sample)
for (i in 2:n_subsamplings) {
    matrix1 <- matrix1 + vegan::rrarefy(eukaryotic_table_transposed, smallest_sample)
}

matrix1 / n_subsamplings -> eukaryotic_table_transposed_rarefied

rm(i, matrix1, eukaryotic_table_transposed)
```

That step can take several minutes.


### rebuild and save the rarefied table

Rarefaction is computationally expensive, so the rarefied table should
be saved:

```{r}
OTU_ids %>%
    bind_cols(eukaryotic_table_transposed_rarefied %>%
              transpose_back_to_tibble) %>%
    remove_empty_OTUs -> eukaryotic_table_rarefied

eukaryotic_table_rarefied %>%
    write_tsv(file = here::here("data", eukaryotic_rarefied_file))
```

Note that there are no empty OTUs after rarefaction.


### sanity check

How many reads per sample in the final table?

```{r}
eukaryotic_table_rarefied %>%
    pivot_longer(cols = -seqid,
                 names_to = "samples",
                 values_to = "reads") %>%
    count(samples, wt = reads, name = "reads") %>%
    select(reads) %>%
    summary()
```

As expected, there are now
`r prettyNum(smallest_sample, scientific=FALSE, big.mark=",")`
 reads in all samples.


Clean up:

```{r}
rm(eukaryotic_rarefied_file, OTU_ids)
```


## NMDS

### local functions

```{r}
. %>%
    vegan::vegdist(., method = "bray") %>%
    vegan::metaMDS(.) -> compute_NMDS_coordinates

. %>%
    vegan::scores(., display = "sites") %>%
    as.data.frame() %>%
    rownames_to_column(var = "samples") %>%
    separate(col = samples,
             into = c("locality", NA),
             sep = 3,
             remove = FALSE) %>%
    separate(col = samples,
             into = c("site_name", "replicate"),
             sep = "_",
             remove = FALSE) %>%
    left_join(metadata, by = "site_name") -> extract_NMDS_coordinates

. %>%
    round(., digits = 4) %>%
    paste("stress: ", ., sep = "") -> format_stress_annotation
```


### computation

Dissimilarity (Bray-Curtis) and ordination (NMDS):

```{r}
set.seed(seed)

eukaryotic_table_transposed_rarefied %>%
    compute_NMDS_coordinates -> occurrences_t.bray.nmds
```

That step can take several minutes.


### prepare results

extract data scores and add stress value:

```{r}
occurrences_t.bray.nmds %>%
    extract_NMDS_coordinates -> data.scores

x_min <- min(data.scores$NMDS1)
x_max <- max(data.scores$NMDS1)
y_max <- max(data.scores$NMDS2)

occurrences_t.bray.nmds$stress %>%
    format_stress_annotation -> stress_annotation
```

Clean up:

```{r}
rm(eukaryotic_table_transposed_rarefied)
```


### variables and functions

```{r}
reds <- RColorBrewer::brewer.pal(6, "Reds")
nmds_title <- "Boye Kinzig rivers 18S V9 (Eukaryota, Bray-Curtis)"

list(
    geom_point(size = 6, shape = 21, colour = "black", stroke = 0.5),
    theme_bw(base_size = 16),
    annotate("text", x = x_max - abs(x_max / 4),
             y = y_max, label = stress_annotation),
    ggtitle(nmds_title)
) -> my_NMDS_theme


plot_per_parameter <- function(parameter_name, legend_title) {

    legend_title %>%
        str_remove("_CRC_class") %>%
        str_replace_all("_", "\n") -> new_legend

    ggplot(data = data.scores,
           aes(x = NMDS1, y = NMDS2, fill = parameter_name)) +
        my_NMDS_theme +
        scale_fill_manual(name = new_legend, values = reds) +
        scale_colour_manual(values = reds)
}
```


### plot (locality)

```{r}
ggplot(data = data.scores,
       aes(x = NMDS1, y = NMDS2, fill = locality)) +
    my_NMDS_theme +
    scale_fill_manual(values = cbPalette) +
    scale_colour_manual(values = cbPalette) -> nmds_plot

nmds_plot
```

No obvious structuration.

Add ellipses representing t-distributions (solid lines) and normal
distributions (dashed lines):

```{r}
nmds_plot +
    stat_ellipse(aes(colour = locality), type = "norm", linetype = 2) +
    stat_ellipse(aes(colour = locality), type = "t", linetype = 1)
```

The t-distribution, also known as Student's t-distribution, is a type
of normal distribution used for smaller sample sizes, where the
variance in the data is unknown.


### plot (all qualitative parameters)

```{r}
data.scores %>%
    select(where(is.factor)) %>%
    map2(.y = names(.),
         ~ plot_per_parameter(.x, .y))
```

Clean up:

```{r}
rm(nmds_plot, occurrences_t.bray.nmds, samples, stress_annotation,
   nmds_title, x_min, x_max, y_max, data.scores)
```


# Parasites

- start from the `eukaryotic_table`,
- keep only parasitic taxa,
- rarefaction,
- NMDS plots

## filtering

For each column whose name starts with taxonomy, for each taxa name in
these columns, discard the probability value, then check if that taxa
name is present in the list of known parasites:

```{r}
eukaryotic_table %>%
    filter(if_any(starts_with("taxonomy"),
                  ~ str_remove(., " *\\([0-9]+\\)") %in% known_parasites)) -> parasitic_table

rm(eukaryotic_table)
```


## list supergroups

```{r}
parasitic_table %>%
    list_supergroups
```

This parasitic dataset is dominated by Rhizaria, Stramenopiles, and
Opisthokonta.


## find smallest sample

```{r}
parasitic_table %>%
    select(-seqid, -starts_with("taxonomy")) %>%
    pivot_longer(cols = everything(),
                 names_to = "samples",
                 values_to = "reads") %>%
    count(samples, wt = reads, name = "reads", sort = TRUE) %>%
    tail(n = 1) %>%
    pull(reads) -> smallest_sample
```


## rarefaction (random subsampling)

### make a transposed table

Remove columns that are not samples:

```{r}
parasitic_table %>%
    select(-seqid, -starts_with("taxonomy")) %>%
    t() -> parasitic_table_transposed

parasitic_table %>%
    select(seqid) -> OTU_ids
```


### randomly subsample the table

Randomly subsample the table, so all samples have the same number of
reads. Repeat the process `r n_subsamplings` times to make sure that
the final profile is as close as possible to the initial
distribution. Use a fix seed to make the process 100% repeatable. That
step can take several minutes to run.

```{r}
set.seed(seed)
matrix1 <- vegan::rrarefy(parasitic_table_transposed, smallest_sample)
for (i in 2:n_subsamplings) {
    matrix1 <- matrix1 + vegan::rrarefy(parasitic_table_transposed, smallest_sample)
}

matrix1 / n_subsamplings -> parasitic_table_transposed_rarefied

rm(i, matrix1, parasitic_table_transposed)
```

That step can take several minutes.


### rebuild and save the rarefied table

Rarefaction is computationally expensive, so the rarefied table should
be saved:

```{r}
OTU_ids %>%
    bind_cols(parasitic_table_transposed_rarefied %>%
              transpose_back_to_tibble) %>%
    remove_empty_OTUs -> parasitic_table_rarefied

parasitic_table_rarefied %>%
    write_tsv(file = here::here("data", parasitic_rarefied_file))
```

Note that there are no empty OTUs after rarefaction.


### sanity check

How many reads per sample in the final table?

```{r}
parasitic_table_rarefied %>%
    pivot_longer(cols = -seqid,
                 names_to = "samples",
                 values_to = "reads") %>%
    count(samples, wt = reads, name = "reads") %>%
    select(reads) %>%
    summary()
```

As expected, there are now
`r prettyNum(smallest_sample, scientific=FALSE, big.mark=",")`
 reads in all samples.


Clean up:

```{r}
rm(parasitic_rarefied_file, OTU_ids)
```


## NMDS

### computation

Dissimilarity (Bray-Curtis) and ordination (NMDS):

```{r}
set.seed(seed)

parasitic_table_transposed_rarefied %>%
    compute_NMDS_coordinates -> occurrences_t.bray.nmds
```


### prepare results

extract data scores and add stress value:

```{r}
occurrences_t.bray.nmds %>%
    extract_NMDS_coordinates -> data.scores

x_min <- min(data.scores$NMDS1)
x_max <- max(data.scores$NMDS1)
y_max <- max(data.scores$NMDS2)

occurrences_t.bray.nmds$stress %>%
    format_stress_annotation -> stress_annotation
```

Clean up:

```{r}
rm(parasitic_table_transposed_rarefied)
```


### variables and functions

```{r}
reds <- RColorBrewer::brewer.pal(6, "Reds")
nmds_title <- "Boye Kinzig rivers 18S V9 (parasites, Bray-Curtis)"
```


### plot (locality)

```{r}
ggplot(data = data.scores,
       aes(x = NMDS1, y = NMDS2, fill = locality)) +
    my_NMDS_theme +
    scale_fill_manual(values = cbPalette) +
    scale_colour_manual(values = cbPalette) -> nmds_plot

nmds_plot
```

No obvious structuration.

Add ellipses representing t-distributions (solid lines) and normal
distributions (dashed lines):

```{r}
nmds_plot +
    stat_ellipse(aes(colour = locality), type = "norm", linetype = 2) +
    stat_ellipse(aes(colour = locality), type = "t", linetype = 1)
```

The t-distribution, also known as Student's t-distribution, is a type
of normal distribution used for smaller sample sizes, where the
variance in the data is unknown.


### plot (all qualitative parameters)

```{r}
data.scores %>%
    select(where(is.factor)) %>%
    map2(.y = names(.),
         ~ plot_per_parameter(.x, .y))
```

Clean up:

```{r}
rm(nmds_plot, occurrences_t.bray.nmds, samples, stress_annotation,
   nmds_title, x_min, x_max, y_max, data.scores)
```



***

```{r}
sessionInfo()
rm(list = ls())
```
